---
word_document:
  fig_height: 7
  fig_width: 10
author: "Justin Gomez and Andrea Mack"
date: ''
output: word_document
title: "Time Series HW 3"
---
__Disclaimer:__ Justin and Andrea are new partners, we have never worked together in one of Dr. Greenwood's classes on an assignment until now.
```{r settings, include = FALSE}
require(xtable)
require(effects)
require(car)#Anova
require(pander)#tables
require(knitr)

knitr::opts_chunk$set(echo = FALSE, comment = NA, warning = FALSE, message = FALSE)
```

## HW 3

You can alone or in pairs that may or may not be people you worked with before. You can discuss it with old partners but should try work as much as possible with new collaborators. 5% bonus if you find someone completely new to work with - that you did not work with on first two assignments.

I mentioned de-seasonalizing of time series, where the seasonal variation is removed from the series to highlight variation at either higher or lower frequencies. There are a variety of techniques for doing this but the simplest is to just subtract the mean for each month from the observations. And the easiest way to find that is using `lm(y~month,data=...)`. 

\newpage

1) For the Bozeman temperature data from HW 1 and 2, estimate a model with month only, subtract its fitted values from the responses (or just extract residuals). Plot the residuals vs the fractional `Year` variable and compare the plot of this result to the plot of the original time series. 

_Comparing the plot of the residuals and the fractional year to the plot of the max monthly temperatures and the fractional year, we can see that the noise has been reduced in the residuals plot, and a slight positive linear trend is observed. Unusual points are also more apparent in the plot of the residuals as they are slightly set apart from the bulk of the data, compared to the plot of the max monthly themperatures with so much noise that extreme points seem to fall in with the rest of the data._

```{r setup, include = FALSE}
setwd("C:/Users/Andrea Mack/Desktop/mack_hub/course_work/Time Series/Homework/HW3/hw3_stat537")
rawbozemandata <- read.csv("rawbozemandata.csv", header = T)

rawd <- rawbozemandata
head(rawd)
rawd$year <- as.numeric(substr(rawd$DATE, 1,4))
rawd$monthf <- as.factor(month.abb[as.numeric(substr(rawd$DATE, 5,6))])
rawd$month <- as.numeric(substr(rawd$DATE, 5,6))
rawd$year.frac <- rawd$year + (rawd$month)/12
rawd$temp <- rawd$MMXT

lm1 <- lm(temp ~ monthf, data = rawd)

lm1.resid <- lm1$residuals

options(show.sigif.stars = FALSE)
```

```{r prob1, fig.height=4, fig.width=6, echo = FALSE}
plot(lm1.resid ~ rawd$year.frac, xlab = "Fractional Year", ylab = "Residuals")
```
  
```{r prob1xa, fig.height=4, fig.width=6, echo = FALSE}

plot(MMXT ~ year.frac, xlab = "Fractional Year", ylab ="MMXT", data = rawd )

```



2) In the de-seasonalized Bozeman temperature data set, re-assess evidence for the linear trend. Compare the result (test statistic, degrees of freedom and size of p-value) of just fitting a linear time trend in these de-seasonalized responses to the result from our original model with a linear year component and a month adjustment (not the quadratic trend model).

_The degrees of freedom change when we are estimating more parameters, and the F statistic changes as well. However, both models resulted in pvalues that were very, very small._

----------------------------------------------------------------
model                     test stat       df                  pvalue
----------                ------         --------             ---------
resid $\sim$ year         Fstat=195.9     $Fstat_{1,1371}$      <0.00001

temp $\sim$ year+monthf   Fstat=1628      $Fstat_{12,1361}$     <0.00001

----------------------------------------------------------------

\label{Deseasonalized Temperature Regressed on Year}
```{r prob2, results = 'asis'}
require(pander)
lm2 <- lm(lm1.resid ~ rawd$year)
options(show.signif.stars = FALSE)
lm2.sum <- summary(lm2)

xtable((lm2.sum))

```

\label{Temperature Regressed on Year and Month}
```{r prob2x, results = 'asis'}
lm3 <- lm(temp ~ year + monthf, data = rawd)
lm3.sum <- summary(lm3)

xtable(lm3.sum)
```

3) I briefly discussed the HADCRUT data set in class. We will consider the HADCRUT4 series of temperature anomalies for the Nothern Hemisphere. The fully up to date data set is available at: http://www.metoffice.gov.uk/hadobs/hadcrut4/data/current/time_series/HadCRUT.4.4.0.0.monthly_nh.txt

```{r prob3, include = FALSE}
setwd("~/Desktop/mack_hub/course_work/Time Series/Homework/HW3/hw3_stat537")
hadcrut <- read.table("hadcrut.txt", header = FALSE)
head(hadcrut)


hadcrut$date <- ts(hadcrut[,1], start = c(1850,01), end = c(2016,07), frequency = 12)
#only use ts if no months/years missing
```

Download the ensemble median monthly northern hemisphere temperature data. We will use the entire time series that is currently available (January 1850 to July 2016). You might want to read http://www.metoffice.gov.uk/hadobs/hadcrut4/data/current/series_format.html for more information on the columns in the data set.

Because the time series is complete over the time frame under consideration, you can use `ts()` to create a time variable instead of messing around with their `Year/Month` variable. 

Make a plot versus time of the ensemble medians and use that as your response variable in the following questions. Discuss trend, seasonality, outliers, and variability.

_Assessing the plot of the ensemble medians versus time, we can see an overall positive trend since January of 1850. The trend has curvature, suggesting a cyclic trend in ensemble medians post 1900. It will be interesting to see if we are currently at a peak in ensemble medians. The resolution of the plot below does not allow us to visally assess seasonality trends. The variability  appears to be larger in the first 50 years of the data set as we can see a range of observations  between about -1.5 and 0.7. In the later years, this spread appears to have decreased. Across the entire 156 year interval, there are a few observations that stand out from the rest of the data as outliers, especially in the first 50 years._

_The next six plots are zoomed in to the first six years. There appears to be seasonality that is visible in the first half of the first four years. Ensemble medians decrease in the first quarter, and then increase in the second quarter of these years. We cannot see seasonality in the third and fourth quarters of these first four years and years five and six do not appear to have a consistent seasonality in them. Other seasonality patterns may or may not emerge in later years_

```{r prob3x, fig.height = 6, fig.width = 8, echo = FALSE}
#col2 are the ensemble medians
par(mfrow=c(1,1))
plot(hadcrut$V2 ~ hadcrut$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")

set.seed(12)
#plot(shuffle(hadcrut$V2) ~ hadcrut$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians", main = "Shuffled")




par(mfrow=c(2,1))
plot(hadcrut[1:12,]$V2 ~ hadcrut[1:12,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")
plot(hadcrut[13:24,]$V2 ~ hadcrut[13:24,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")
par(mfrow = c(2,1))
plot(hadcrut[25:36,]$V2 ~ hadcrut[25:36,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")
plot(hadcrut[37:48,]$V2 ~ hadcrut[37:48,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")
plot(hadcrut[49:60,]$V2 ~ hadcrut[49:60,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")
plot(hadcrut[61:72,]$V2 ~ hadcrut[61:72,]$date, xlab = "Months Since January 1850", ylab = "Ensemble Medians")


#an ensemble data set in which the 100 constituent
#ensemble members sample the distribution of likely surface temperature anomalies given our
#current understanding of these uncertainties
#Sea Surface Temperature anomalies in degrees Celsius, or "SST anomalies" for short, are how much #temperatures depart from what is normal for that time of year.
```
4) Our main focus with these data will be on estimating the long-term trend, starting with polynomial trend models. But first, check for seasonality in a model that accounts for a linear time trend. Use our previous fractional year for the trend. Report an `effects` plot and a test for the month model component.

$H_{o}$: all Bmonth_j = 0

$H_{a}$: at least one Bmonth_j \neq$ 0

_Based on an F-stat of 6.74 on 11 and 1986 degrees of freedom, using type II sums of squares, with an associated pvalue of < 0.001, there is strong evidence that the true mean of the median ensemble changes by month after accounting for fractional date. Examining the effect plot for the monthly factor, we see seasonality in the ensemble medians with average ensemble medians decreasing in January to March, increasing until August, and then decreasing until November, where there is a slight increase again in December._ 


```{r prob4, results = 'asis'}

#%%__we can see there is very little overlap in the confidence intervals for the median ensemble #%%estimate for many pairs of months, indicating time has an affect on the monthly median ensemble #changes. WE SHOULD DISCUSS THIS INTERPRETATION__

hadcrut$monthf <- as.factor(as.numeric(substr(hadcrut$V1,6,7)))
hadcrut$time.date <- as.vector(time(hadcrut$date))
lm4 <- lm(V2 ~ time.date + monthf, data = hadcrut)#don't need monthf as explanatory bc month already induced in time??

plot(allEffects(lm4))

lm4.anova <- Anova(lm4, type = "II")
(xtable(lm4.anova))

```
Note: when you use `time()` to generate the `Year` variable from a time series object it retains some time series object information that can cause conflicts later on. Create a new variable in your data.frame that uses something like `as.vector(time(tsdataname))`.

5) Check the residuals versus fitted values for any evidence of nonlinearity in the residuals vs fitted that was missed by the model with a linear trend and month component. Also note any potential issues with the constant variance assumption.

_Viewing the residuals vs. fitted values plot below, there appears to be a quadratic relationship between the fitted values and residuals that was not captured by the model. For fitted values less than -0.3, there appears to be more variation in the residuals than for fitted values above -0.3, indicating the constant variance assumption may be violated. The model is less accurately predicting median ensembles when the fitted median ensemble was below -0.3, which means that we may have left out a variable or structure that explains when median ensemble was lower than expected._

```{r prob5}
par(mfrow = c(1,1))
plot(lm4,which=1)


```

6) You can add higher order polynomial terms to models using `x1+I(x1^2)+I(x1^3)`... or using the `poly` function, such as `poly(x1,3,raw=T)` for a cubic polynomial that includes the linear and quadratic components (we want this!). The `raw=T` keeps the variables in their raw or input format. Estimate the same model but now using polynomial trends that steps up from linear (poly(time,1,raw=T)) and stop when you get a failure to estimate a part of the model. Briefly discuss what happened.

When I got to the fifth degree polynomial, the estimate could not be computed. WHY? My guess is that it has something to do with the seasonality of the time.data variable, or it be perfectly collinear with another variable.

_We can see that the estimate for the fifth degree term in the model cannot be computed, while all lower order terms could be. Examining the estimates themselves, we can see the intercept is a large number, as is the estimate for the first degree term. The estimate for the second degree term is relatively small compared to the first two estimates, with the third and fourth degree terms being even smaller. One reason the fifth degree term returns NAs could be that it is almost nearly collinear with one or a combination of the other predictors, making the coefficient for the fifth degree term not estimable._

_Partial output from the model with a 5th degree polynomial term in it. You will notice that the output does not include a line for the 5th degree term. This is a formatting issue we were having and could not get resolved. The actual output for the 5th degree term returns NA's, which means that that coefficient could not be estimated._

```{r prob6, include = FALSE}
lm.d1 <- lm(V2~poly(time.date,1,raw=TRUE) + monthf, data = hadcrut)
summary(lm.d1)
test<-lm(V2~monthf,data=hadcrut)
resids1<-test$residuals
par(mfrow=c(1,1))

plot(resids1^5~hadcrut$time.date,xlim = c(1850,1900), ylim = c(-0.5,0.5))
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time)), add = TRUE, col = "yellow")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^2, add = TRUE, col = "purple")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^3, add = TRUE, col = "red")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^4, add = TRUE, col = "blue")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^5, add = TRUE, col = "green")

plot.window(xlim = c(1900,1950), ylim = c(-3,3))
plot(resids1^5~hadcrut$time.date,xlim = c(1900,1950), ylim = c(-0.075,0.075))
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^3, add = TRUE, col = "red")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^4, add = TRUE, col = "blue")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^5, add = TRUE, col = "green")

plot(resids1^5~hadcrut$time.date,xlim = c(1950,2016), ylim = c(-0.5,0.5))
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time)), add = TRUE, col = "yellow")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^2, add = TRUE, col = "purple")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^3, add = TRUE, col = "red")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^4, add = TRUE, col = "blue")
curve((1/sd(hadcrut$time))*((x-mean(hadcrut$time))/sd(hadcrut$time))^5, add = TRUE, col = "green")

lm.d <- NULL
for(i in 1:25){
lm.d[i] <- lm(V2~poly(time.date,i,raw=TRUE) + monthf, data = hadcrut)
}

#s4 <- summary(lm.d[[4]])
#lm.d[5]
#
```
  
  


```{r k, results = 'asis'}
#another way, I was getting errors in 7 with the for loop
polyfunc<-function(i) {
  poly.lm<-lm(V2~poly(time.date,i,raw=TRUE) + monthf, data = hadcrut)
  return (summary(poly.lm))
}
#xtable(polyfunc(4))
require(knitr)
kable(polyfunc(5)$coeff)
```

------------------------------------------------------------------------------------------------
|                                |      Estimate|   Std. Error|     t value| Pr(>&#124;t&#124;)|
|:-------------------------------|-------------:|------------:|-----------:|------------------:|
|(Intercept)                     |  2.146402e+05| 2.102369e+04|  10.2094424|          0.0000000|
|poly(time.date, i, raw = TRUE)1 | -4.458441e+02| 4.352932e+01| -10.2423869|          0.0000000|
|poly(time.date, i, raw = TRUE)2 |  3.472438e-01| 3.378960e-02|  10.2766371|          0.0000000|
|poly(time.date, i, raw = TRUE)3 | -1.202000e-04| 1.170000e-05| -10.3123040|          0.0000000|
|poly(time.date, i, raw = TRUE)4 |  0.000000e+00| 0.000000e+00|  10.3494916|          0.0000000|
|monthf2                         | -3.290620e-02| 2.703940e-02|  -1.2169716|          0.2237598|
|monthf3                         | -1.073839e-01| 2.703940e-02|  -3.9713798|          0.0000740|
|monthf4                         | -1.649880e-02| 2.703950e-02|  -0.6101762|          0.5418150|
|monthf5                         | -5.424800e-03| 2.703950e-02|  -0.2006249|          0.8410124|
|monthf6                         |  5.012570e-02| 2.703960e-02|   1.8537911|          0.0639175|
|monthf7                         |  7.245800e-02| 2.703960e-02|   2.6796976|          0.0074298|
|monthf8                         |  7.885290e-02| 2.708040e-02|   2.9118085|          0.0036336|
|monthf9                         |  4.594480e-02| 2.708040e-02|   1.6966074|          0.0899278|
|monthf10                        |  3.252210e-02| 2.708040e-02|   1.2009454|          0.2299158|
|monthf11                        | -6.024050e-02| 2.708050e-02|  -2.2244985|          0.0262271|
|monthf12                        | -5.592130e-02| 2.708050e-02|  -2.0650010|          0.0390526|
------------------------------------------------------------------------------------------------

7) If we center or, even better, make the polynomial functions orthogonal to one another, we can avoid the issue in the previous question. Use `poly(x1,?,raw=F)` and step up the polynomial order for time until the p-value for the last coefficient (use `summary()`) is "large", reporting the single test result for each step in the building process. Then drop back one order and re-fit the model. Report the `effects` plot of the resulting model and describe the estimated trend. Note: aside from access to orthogonal polynomials the `poly` function interfaces with `Anova` and the `effects` package really nicely.

_The notation we are using assumes the intercept is B0, the linear term is B1, the quadratic term is B2, etc. The results from testing the largest polynomial term in each model are given below. The model with a 10th degree polynomial has a pvalue of 0.17 that provides no evidence that the coefficient of the 10th degree polynomial is different from zero, after considering the information in first 9 polynomial terms and the month factor terms._

```{r prob 7, results = 'asis'}
#hadcrut$time.date.scale <- c(scale(hadcrut$time.date))
a <- seq(1:15)
#we are not supposed to scale these, by using raw = FALSE, r is automatically fitting orthogonal polynomials

s <- lapply(a, function(x){summary(lm(V2~poly(time.date,x,raw=FALSE) + monthf, data = hadcrut))})

a2 <- seq(2:15)
test <- NULL
for(i in a2){
cat("Ho:", "B",i-1, " = 0", "Ha", "B", i-1, "$\\neq 0$", "tstat = ", s[[i]]$coefficients[i,3], "df = ", 1987-i, "pvalue = ", s[[i]]$coefficients[i,4],"\n", sep = "\n")
}
```
 
 
```{r prob7x, fig.algin = 'center'}
#orthofunc<-function(i) {
  #ortho.lm<-lm(V2~poly(time.date,i,raw=FALSE) + monthf,data=hadcrut)
  #return (summary(ortho.lm))
#}
###orthofunc(10) #"large p value"

final.lm<-lm(V2~poly(time.date,9,raw=FALSE) + monthf,data=hadcrut)
plot(allEffects(final.lm))

```


8) Check the diagnostic plots from your final model. Does anything improve from the first version. Also plot the residuals vs time and compare that plot to residuals vs fitted.  

_The residuals versus fitted plot improved in terms of linearity. To assume that the residuals are linearly related to the response is now reasonable. However, other assumptions are now largely violated. In the residuals vs. fitted plot we see the variation much larger for smaller fitted values than for larger fitted values. While the sample size is large enough to not be too concerned with violations to normality, the residuals are farther from the normal QQ line with the 9th degree polynomial than with the linear model. We very clearly see long tails in the nomral QQ plot with the 9th degree polynomial. Observations 517 and 133 appear to be outliers._

_Comparing the plots of the residuals from the 9th degree ploynomial to those of the linear model, we see that the curvature is nearly gone, but there is still higher variation in the earlier years._

```{r prob8, fig.align = 'center'}
par(mfrow=c(2,2))
plot(final.lm)
par(mfrow=c(1,1))
plot(final.lm$residuals~hadcrut$time.date, xlab = "Time", ylab = "9th degree poly residuals")

plot(final.lm$residuals ~ final.lm$fitted.values, xlab = "9th degree poly fitted", ylab = "9th degree poly residuals")
```


9) Run the following code so I can see what version of R you are now using:

### Documenting R version 

```{r}getRversion()
```

__R Code__

1)
```{r settings, eval = FALSE, echo = TRUE}
```

```{r setup, eval = FALSE, echo = TRUE}
```

```{r prob1, eval = FALSE, echo = TRUE}
```

```{r prob1xa, eval = FALSE, echo = TRUE}
```

2)
```{r prob2, eval = FALSE, echo = TRUE}
```

```{r prob2x, eval = FALSE, echo = TRUE}
```

3) 
```{r prob3, eval = FALSE, echo = TRUE}
```

```{r prob3x, eval = FALSE, echo = TRUE}
```

4)
```{r prob4, eval = FALSE, echo = TRUE}
```


5)
```{r prob5, eval = FALSE, echo = TRUE}
```

6)
```{r prob6, eval = FALSE, echo = TRUE}
```

other
```{r k, eval = FALSE, echo = TRUE}
```

7)
```{r prob7, eval = FALSE, echo = TRUE}
```

```{r prob7x, eval = FALSE, echo = TRUE}
```

8)
```{r prob8, eval = FALSE, echo = TRUE}
```










